---
title: "Health dimensions by regimes"
author: "Maria Potterf"
date: "September 17, 2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Forest health indicators for SIMO

0 = absent
1 = present

#### Shelter
- **even/uneven-aged** forest: even-aged= if 80% of the volume is located in age classes 20 years apart
- SIMO *quality* - even/uneven age: even-age = if 80% of volume is located within range of 20 years
- indicator presence/absence of the **layers of less then 5 m**
- SIMO: *Layers_lt_2m*    - number of layers under 2 m
- SIMO: *Layers_lt_2m_N*  - number of stems/ha under 2 m
- SIMO: *Layers_lt_5m*    - number of layers under 5 m   
- SIMO: *Layers_lt_5m_N*  - number of stems/ha under 5 m

#### Serenity
- if distance > 400m from road at 70km/h max speed
- variable derived from spatial data


#### Diversity
- SIMO: *Volume* for individual species;
- calculate the % of volume by individual tree species 

**Tree species**

- *V_pine*
- *V_spruce*                 
- *V_Birch_1*
- *V_Birch_2*
- *V_populus*
- *V_Alnus_incana*
- *V_Alnus_glutinosa*
- *V_o_coniferous*           
- *V_o_decidious*                            
- If > 70% of volume is created by one species, then not a diverse forest

#### Nature
- natural if mean age > 100 years
- SIMO: *AGE_vol* - age weighthed by volume
- SIMO: *AGE_ba*  - age weighthed by basal area

#### Cohesion
- if stand area > 1 ha

## How to combine dimensions? 

- **Edges/glades**:    shelter, nature, diversity
- **Deep forest**:     nature, shelter, serenity
- **Spacious forest**: cohesion, serenity, nature


### Forest management regimes:

Select the most diverse management regimes: 

- **BAU** - business as usual: final cut, including thinnings
- **BAU_GTR** - business as usual, with retained trees after final harvest (Green Tree Retention)
- **CCF** - continuous cover forestry, no final cut
- **SA**  - set aside, no harvest


``` {r read-libs, include=FALSE}
rm(list = ls())

#setwd("C:/MyTemp/myGitLab/windDamage")

# Read libraries
library(ggplot2)
library(dplyr)
library(tidyr)
library(sf)
library(rgdal)
library(dplyr)
library(spData)
library(sf)
library(purrr)

```

### Forest restoration potential by management regimes


```{r get-data, include=FALSE}

inPath = "C:/MyTemp/myGitLab/bayesianESS/build_ess_models_bayesian"
inFolder = "input"

# Read data
df <- data.table::fread(paste(inPath, 
                              inFolder, 
                              "RCP85_p_301-600_l3.csv",  
                              sep = "/"),  
                        data.table=FALSE, stringsAsFactors = FALSE)

```

```{r subset-data, include=FALSE}

# Filter data first
my_regimes = c("SA_DWextract", "BAU",  "CCF", "BAUwT_GTR")
my_cols = c("id",
           "year",
           'AREA',
           "branching_group", 
           "Age", 
           "BA", 
            "V", 
           #"Harvested_V_log",
           #"Harvested_V_pulp", 
          #"PEAT",
            "H_dom"  ,
            # "D_gm", 
            #     "COMBINED_HSI"  
            "AGE_vol", 
            #"AGE_ba", 
            "V_pine",
           "V_spruce",
           "V_Birch_1",
           "V_Birch_2",
           "V_populus",
           "V_Alnus_incana",
           "V_Alnus_glutinosa",
           "V_o_coniferous",
           "V_o_decidious",
           "quality",
           "Layers_lt_2m",
           "Layers_lt_5m",
           "Layers_lt_2m_N",
           "Layers_lt_5m_N",
            "name", 
           "regime")


df <- df %>% 
  filter(regime %in% my_regimes) %>% 
  dplyr::select(all_of(my_cols))


# Recode regimes to have nicer regime names
df$regime <- dplyr::recode_factor(as.factor(df$regime),
                                  BAU = "BAU",
                BAUwT_GTR = 'BAU_GTR',
                CCF = 'CCF',
                SA_DWextract = 'SA')

df$regime <- as.character(df$regime) 

```


SIMO: **quality** variable contains many NA values: these follows recent harvest, e.g. H_dom (tree height) is NA or very low. Consider these as 'no forest', or keep as NA

**Calculate the % of species by volume**

if any of the species has more that > 70% of volume, that the stands is not mixed = not diverse


```{r get-metrics, include=FALSE}
# Calculate the volume proportion by species, V = total volume
# count how many species are present = richness
# e.g. count number of species (= columns) that volume is not NA

library(dplyr)

cols_species = c(
  'V_pine',
  'V_spruce',
  'V_Birch_1',
  'V_Birch_2',
  'V_populus',
  'V_Alnus_incana',
  'V_Alnus_glutinosa',
  'V_o_coniferous',
  'V_o_decidious'
)



# Count number or columns that have some volume (exist) in a stand
df$richness <- rowSums( !is.na( select(df, all_of(cols_species))))


# Get proportion of volume per species from total volume: `V`
# select all columns containing species and divide it bu total volume
# reclassify based on rule: diverse if <70% of volume is made by one species
df.V.prop = select(df, all_of(cols_species))/df$V*100


# Classify as mixed/no mixed if any of species column have less then 70% of volume in single species 
df$q_mixed <- 
  df.V.prop %>%
   rowwise() %>%
  mutate(q_mixed = case_when(all(is.na(c_across())) ~ NA_character_, 
                            any(c_across() >= 70) ~ 'uniform', 
                            TRUE ~ 'mixed')) %>% 
    pull(q_mixed)


```



#### Intial conditions



```{r hist-age2016, echo = F, fig.dim = c(3, 2.5), include=TRUE, message=FALSE}

df %>% 
  filter(year == 2016) %>%
  ggplot(aes(x = AGE_vol)) +
  geom_histogram(fill = 'lightgrey',
                 color = 'black') +
  ylab("Age [years]") +
 # scale_color_viridis_d() +
  xlab("year") +
    theme_bw() +
  ggtitle("Age distribution [2016]") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, 
                                   vjust = 1, 
                                   hjust=1))  





```


```{r initial, echo = F, fig.dim = c(7,3), include=TRUE, message=FALSE}

df %>% 
  #filter(year == 2016) %>%
  ggplot(aes(y = AGE_vol,
             x = year, #as.factor(year),
             group = year,
             fill = regime)) +
  geom_boxplot(   outlier.size = 0.5)+
  facet_grid(.~regime) + 
  scale_fill_viridis_d() + 
  ylab("Age [years]") +
  scale_color_viridis_d() +
  xlab("year") +
  theme_bw() +
    theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 45, 
                                   vjust = 1, 
                                   hjust=1)) 




```




### Dimensions

#### Shelter:

- if quality is uneven-aged and have a any layers in the height of 5m:
- 0 - no shelter
- 1 - yes shelter


```{r shelter, echo=FALSE, fig.dim = c(7, 2.5), include=FALSE}

# Classify sheltering

df <- 
  df %>% 
  mutate(shelter = case_when(
    (quality == "uneven-aged" & "Layers_lt_5m" >= 1) ~ 1, 
    TRUE ~0 )) 


```


```{r plot-shelter, echo = F, fig.dim = c(7, 2.5)}

df %>% 
  group_by(year, regime, shelter) %>% 
  tally() %>%  # get counts
  ggplot(aes(x = year,
             y = n,
             color = regime)) +
  geom_line(lwd = 1) +
  facet_grid(.~shelter) + 
  theme_bw() +
  ylab("# of stands") +
  scale_color_viridis_d() +
  ggtitle("Number of stands with shelter quality")

```


#### Serenity: 

- distance from major roads
- Daniel show road network over the forest stand geometry?
- 0 - within 400 m from the road of 70km/ha max speed
- 1 - more then 400 m away (edge-to-edge or centroid distance??)

#### Diversity:  

- diverse if species are mixed (e.g. >70% of the species volume is created by more then one species, one species has at most 70% of total volume)
- 0 - no diverse
- 1 - yes diverse



```{r diversity, echo=FALSE, fig.dim = c(7, 2.5), include=T}

# Classify diversity

df <- 
  df %>% 
  mutate(divers = case_when(
    q_mixed == "mixed" ~ 1, 
    TRUE~ 0)) 

```



```{r plot-diversity, echo=FALSE, fig.dim = c(7, 2.5), include=T}

df %>% 
  group_by(year, regime, divers) %>% 
  tally() %>%  # get counts
  ggplot(aes(x = year,
             y = n,
             color = regime)) +
  geom_line(lwd = 1) +
  facet_grid(.~divers) + 
  scale_color_viridis_d() +
  theme_bw() +
  ylab("# of stands") +
  ggtitle("Number of stands with diversity quality")

```

```{r plot-richness, echo = F, include=F, fig.dim = c(7, 2.5)}

df %>%
  #group_by(year, regime, richness) %>%
  #tally() %>%  # get counts
  ggplot(aes(x = richness,
             y = V,
             color = regime)) +
  geom_jitter(alpha = 0.5) 
```

#### Naturalness: 

- age > 100 years old

Mean age of all stands is  `r round(mean(df$AGE_vol), 1)` years old. Distribution by regimes: 

```{r hist-age, fig.dim = c(7, 2.5), include=FALSE, message=FALSE}

df <- df %>% 
  mutate(natural = case_when(
    AGE_vol >= 100 ~ 1, 
    TRUE~ 0))

```


```{r plot-natural, echo = F, fig.dim = c(7, 2.5), message=FALSE}

df %>% 
  group_by(year, regime, natural) %>% 
  tally() %>%  # get counts
  ggplot(aes(x = year,
             y = n,
             color = regime)) +
  geom_line(lwd = 1) +
  facet_grid(.~natural) + 
  theme_bw() +
  scale_color_viridis_d() +
  ylab("# of stands") +
  ggtitle("Number of stands with naturalness quality")

```


#### Cohesion: AREA is missing from this data!! 
(made from NFI plots, we need to add geometry layer instead)

- stands over 1 ha are cohesive

Mean stand area is `r round(mean(df$AREA, na.rm = T), 3)` ha. 

```{r cohesion, echo=FALSE, fig.dim = c(7, 2.5), include=FALSE}

# Classify cohesion

df <- 
  df %>% 
  mutate(cohesive = case_when(
    AREA >= 1 ~ 1, 
    TRUE~ 0)) 

```


# Combine dimensions into forest categories:

- **multiplication** = needs to fulfill all qualities 
edges = shelter * nature * diversity

- **addition** = needs to fulfill at least one quality
edges = shelter + nature + diversity


```{r edge, echo=FALSE,  include=FALSE}

df <- df %>% 
  mutate(edge_multi = shelter *divers* natural,
         edge_add =  shelter +divers+ natural)

```

```{r edge-plot, echo=FALSE, fig.dim = c(7, 2.5), include=T}

df %>% 
  group_by(year, regime, edge_multi) %>% 
  tally() %>%  # get counts
  ggplot(aes(x = year,
             y = n,
             color = regime)) +
  geom_line(lwd = 1) +
  facet_grid(.~edge_multi) + 
  theme_bw() +
  scale_color_viridis_d() +
  ylab("# of stands") +
  ggtitle("Number of stands with edge quality = multi")
  
  
df %>% 
  group_by(year, regime, edge_add) %>% 
  tally() %>%  # get counts
  ggplot(aes(x = year,
             y = n,
             color = regime)) +
  geom_line(lwd = 1) +
  facet_grid(.~edge_add) + 
  theme_bw() +
  scale_color_viridis_d() +
  ylab("# of stands") +
  ggtitle("Number of stands with edge quality = addit")
 
```
